Different rules for binocular combination of luminance flicker in cortical and subcortical pathways

### 2023-06-26 manuscript.Rmd
Added references and other minor tweaks

The pupillary light reflex is an automatic constriction of the iris sphincter muscles in response to increases in light levels, which causes the pupil to shrink [@McDougal2008]. There is a clear binocular component to this reflex, as stimulation of one eye still causes constriction of the other eye's pupil (termed the consensual response; @Wyatt1981). Importantly, the neuroanatomical pathway involved completely bypasses the canonical cortical pathway (retina to V1), instead involving a network of subcortical nuclei, including the Pretectal Olivary nucleus, Superior Cervical ganglion, and Edinger-Westphal nucleus [@Angee2021; @Mathot2018; @McDougal2008; @Wang2015]. To account for the consensual response, these brain regions must combine information from the left and right eyes [@Doesschate1967], yet the computation that achieves this is achieved is unclear. The pupil response can be modulated by periodic changes in luminance, and is temporally low-pass [@Barrionuevo2014; @Spitschan2014], most likely due to the mechanical limitations of the iris sphincter and dilator muscles.

@@

We fitted a computational model to the data from Experiments 1 & 3 using a hierarchical Bayesian approach. The model behaviour is displayed in Figure \@ref(fig:modelfigure)e-h, with empirical data replotted in Figure \@ref(fig:modelfigure)a-d for comparison. In general, the model captures the key characteristics of the empirical data, with group-level parameter estimates provided in Table \@ref(tab:paramtable). We were particularly interested in comparing the weight of interocular suppression across data sets. We therefore plot the posterior distributions for this parameter for all four data sets (see Figure \@ref(fig:modelfigure)i). The key finding is that the pupillometry results (green distribution) display a much greater weight of interocular suppression compared with the other data sets (grey, purple and yellow distributions). There is no overlap between the pupillometry distribution and any of the other three. All four distributions are also meaningfully below a weight of 1 -- the value that previous work using grating stimuli would predict [@Meese2006; @Baker2017], and the peak location of our prior distribution (black curve). These results offer an explanation of the empirical data: the strong interocular suppression for the pupillometry data is consistent with the weak binocular facilitation, and measurable dichoptic masking observed using that method. The weaker suppression for the other experiments is consistent with the near-linear binocular facilitation effects, and absent dichoptic masking.

@@

for (cond in 1:3){
  points(contrastsdB,abs(cleanmeansP[cond,]), pch = 20+cond, col='black', bg=colpal[cond], cex=1.6, lwd=3)   # draw the data points themselves
}  
cond <- 6
  points(contrastsdB,abs(cleanmeansP[cond,]), pch = 25, col='black', bg=colpal[cond], cex=1.6, lwd=3)

legend(21,0.03,c('Monocular','Binocular','Dichoptic','Dichoptic cross'),pch=c(21:23,25),pt.bg=colpal[c(1:3,6)],pt.lwd=3,pt.cex=1.6,box.lwd=2)
text(17,0.028,'(a)',adj=0.5,cex=2.5)

@@

for (cond in 1:3){
  points(contrastsdB,abs(cleanmeansEav[cond,]), pch = 20+cond, col='black', bg=colpal[cond], cex=1.6, lwd=3)   # draw the data points themselves
}  
cond <- 6
  points(contrastsdB,abs(cleanmeansEav[cond,]), pch = 25, col='black', bg=colpal[cond], cex=1.6, lwd=3)   # draw the data points

@@

for (cond in 1:3){
  points(contrastsdB,abs(cleanmeansE2av[cond,]), pch = 20+cond, col='black', bg=colpal[cond], cex=1.6, lwd=3)   # draw the data points themselves
}  
cond <- 6
  points(contrastsdB,abs(cleanmeansE2av[cond,]), pch = 25, col='black', bg=colpal[cond], cex=1.6, lwd=3)

@@

```{r modelfigure, fig.cap="Summary of computational modelling. Panels (a-d) show empirical data from key conditions, replotted from earlier figures for the pupillometry (a), first harmonic EEG responses (b), second harmonic EEG responses (c) and contrast matching (d) experiments. Panels (e-h) show model behaviour for the same conditions, generated using the median group-level parameter values.  Panel (i) shows the posterior probability distributions of the interocular suppression parameter for each of the four model fits. The pupillometry distribution (green) is centred about a substantially higher suppressive weight than for the other data types (note the logarithmic x-axis). The black curve shows the (scaled) prior distribution for the weight parameter.", fig.align="center", echo=FALSE}

@@

We conducted a conceptual replication of Experiment 1 using an alternative system. A pair of Spectra Tune Lab multiprimary devices (LEDmotive Technologies LLC, Barcelona, Spain) were coupled to a binocular headset using liquid light guides. The light was imaged onto a circular diffuser for each eye (field size 30 deg) with the central 8 degrees masked off using a black occluder. Therefore the replication experiment involved peripheral stimulation, unlike the main experiment which stimulated the central ~4 degrees of the visual field. All conditions were otherwise as described for the main experiment, and we tested 12 participants in total.

>>>>

The pupillary light reflex is an automatic constriction of the iris sphincter muscles in response to increases in light levels, which causes the pupil to shrink [@McDougal2008]. There is a clear binocular component to this reflex, as stimulation of one eye still causes constriction of the other eye's pupil (termed the consensual response; @Wyatt1981). Importantly, the neuroanatomical pathway involved completely bypasses the canonical cortical pathway (retina to V1), instead involving a network of subcortical nuclei, including the Pretectal Olivary nucleus, Superior Cervical ganglion, and Edinger-Westphal nucleus [@Angee2021; @Mathot2018; @McDougal2008; @Wang2015]. To account for the consensual response, these brain regions must combine information from the left and right eyes [@Doesschate1967], yet the computation that achieves this is unclear. The pupil response can be modulated by periodic changes in luminance, and is temporally low-pass [@Barrionuevo2014; @Spitschan2014], most likely due to the mechanical limitations of the iris sphincter and dilator muscles [@Privitera2006].

@@

We fitted a computational model to the data from Experiments 1 & 3 using a hierarchical Bayesian approach. The model behaviour is displayed in Figure \@ref(fig:modelfigure)a-d, with empirical data superimposed for comparison. In general, the model captures the key characteristics of the empirical data, with group-level parameter estimates provided in Table \@ref(tab:paramtable). We were particularly interested in comparing the weight of interocular suppression across data sets. We therefore plot the posterior distributions for this parameter for all four data sets (see Figure \@ref(fig:modelfigure)e). The key finding is that the pupillometry results (green distribution) display a much greater weight of interocular suppression compared with the other data sets (grey, purple and yellow distributions). There is no overlap between the pupillometry distribution and any of the other three. All four distributions are also meaningfully below a weight of 1 -- the value that previous work using grating stimuli would predict [@Meese2006; @Baker2017], and the peak location of our prior distribution (black curve). These results offer an explanation of the empirical data: the strong interocular suppression for the pupillometry data is consistent with the weak binocular facilitation, and measurable dichoptic masking observed using that method. The weaker suppression for the other experiments is consistent with the near-linear binocular facilitation effects, and absent dichoptic masking.

@@

for (cond in 1:3){
  points(contrastsdB,abs(cleanmeansP[cond,]), pch = 20+cond, col=colpal[cond], bg=colpal[cond], cex=1.6, lwd=3)   # draw the data points themselves
}  
cond <- 6
  points(contrastsdB,abs(cleanmeansP[cond,]), pch = 25, col=colpal[cond], bg=colpal[cond], cex=1.6, lwd=3)

legend(21,0.03,c('Monocular','Binocular','Dichoptic','Dichoptic cross'),pch=c(21:23,25),pt.bg=colpal[c(1:3,6)],lwd=3,lty=1,col=colpal[c(1:3,6)],pt.lwd=3,pt.cex=1.6,box.lwd=2)
text(17,0.028,'(a)',adj=0.5,cex=2.5)

@@

for (cond in 1:3){
  points(contrastsdB,abs(cleanmeansEav[cond,]), pch = 20+cond, col=colpal[cond], bg=colpal[cond], cex=1.6, lwd=3)   # draw the data points themselves
}  
cond <- 6
  points(contrastsdB,abs(cleanmeansEav[cond,]), pch = 25, col=colpal[cond], bg=colpal[cond], cex=1.6, lwd=3)   # draw the data points

@@

for (cond in 1:3){
  points(contrastsdB,abs(cleanmeansE2av[cond,]), pch = 20+cond, col=colpal[cond], bg=colpal[cond], cex=1.6, lwd=3)   # draw the data points themselves
}  
cond <- 6
  points(contrastsdB,abs(cleanmeansE2av[cond,]), pch = 25, col=colpal[cond], bg=colpal[cond], cex=1.6, lwd=3)

@@

```{r modelfigure, fig.cap="Summary of computational modelling. Panels (a-d) show empirical data from key conditions, replotted from earlier figures for the pupillometry (a), first harmonic EEG responses (b), second harmonic EEG responses (c) and contrast matching (d) experiments, with curves showing model behaviour generated using the median group-level parameter values.  Panel (e) shows the posterior probability distributions of the interocular suppression parameter for each of the four model fits. The pupillometry distribution (green) is centred about a substantially higher suppressive weight than for the other data types (note the logarithmic x-axis). The black curve shows the (scaled) prior distribution for the weight parameter.", fig.align="center", echo=FALSE}

@@

We conducted a conceptual replication of Experiment 1 using an alternative system of hardware and software [@Martin2022; @Martin2023]. A pair of Spectra Tune Lab multiprimary devices (LEDmotive Technologies LLC, Barcelona, Spain) were coupled to a binocular headset using liquid light guides. The light was imaged onto a circular diffuser for each eye (field size 30 deg) with the central 8 degrees masked off using a black occluder. Therefore the replication experiment involved peripheral stimulation, unlike the main experiment which stimulated the central ~4 degrees of the visual field. All conditions were otherwise as described for the main experiment, and we tested 12 participants in total.

### 2023-06-26 manuscript.tex
Added references and other minor tweaks

The pupillary light reflex is an automatic constriction of the iris sphincter muscles in response to increases in light levels, which causes the pupil to shrink (\protect\hyperlink{ref-McDougal2008}{McDougal and Gamlin, 2008}). There is a clear binocular component to this reflex, as stimulation of one eye still causes constriction of the other eye's pupil (termed the consensual response; Wyatt and Musselman (\protect\hyperlink{ref-Wyatt1981}{1981})). Importantly, the neuroanatomical pathway involved completely bypasses the canonical cortical pathway (retina to V1), instead involving a network of subcortical nuclei, including the Pretectal Olivary nucleus, Superior Cervical ganglion, and Edinger-Westphal nucleus (\protect\hyperlink{ref-Angee2021}{Angée et al., 2021}; \protect\hyperlink{ref-Mathot2018}{Mathôt, 2018}; \protect\hyperlink{ref-McDougal2008}{McDougal and Gamlin, 2008}; \protect\hyperlink{ref-Wang2015}{Wang and Munoz, 2015}). To account for the consensual response, these brain regions must combine information from the left and right eyes (\protect\hyperlink{ref-Doesschate1967}{Doesschate and Alpern, 1967}), yet the computation that achieves this is achieved is unclear. The pupil response can be modulated by periodic changes in luminance, and is temporally low-pass (\protect\hyperlink{ref-Barrionuevo2014}{Barrionuevo et al., 2014}; \protect\hyperlink{ref-Spitschan2014}{Spitschan et al., 2014}), most likely due to the mechanical limitations of the iris sphincter and dilator muscles.

@@

We fitted a computational model to the data from Experiments 1 \& 3 using a hierarchical Bayesian approach. The model behaviour is displayed in Figure \ref{fig:modelfigure}e-h, with empirical data replotted in Figure \ref{fig:modelfigure}a-d for comparison. In general, the model captures the key characteristics of the empirical data, with group-level parameter estimates provided in Table \ref{tab:paramtable}. We were particularly interested in comparing the weight of interocular suppression across data sets. We therefore plot the posterior distributions for this parameter for all four data sets (see Figure \ref{fig:modelfigure}i). The key finding is that the pupillometry results (green distribution) display a much greater weight of interocular suppression compared with the other data sets (grey, purple and yellow distributions). There is no overlap between the pupillometry distribution and any of the other three. All four distributions are also meaningfully below a weight of 1 -- the value that previous work using grating stimuli would predict (\protect\hyperlink{ref-Baker2017}{Baker and Wade, 2017}; \protect\hyperlink{ref-Meese2006}{Meese et al., 2006}), and the peak location of our prior distribution (black curve). These results offer an explanation of the empirical data: the strong interocular suppression for the pupillometry data is consistent with the weak binocular facilitation, and measurable dichoptic masking observed using that method. The weaker suppression for the other experiments is consistent with the near-linear binocular facilitation effects, and absent dichoptic masking.

@@

\caption{Summary of computational modelling. Panels (a-d) show empirical data from key conditions, replotted from earlier figures for the pupillometry (a), first harmonic EEG responses (b), second harmonic EEG responses (c) and contrast matching (d) experiments. Panels (e-h) show model behaviour for the same conditions, generated using the median group-level parameter values.  Panel (i) shows the posterior probability distributions of the interocular suppression parameter for each of the four model fits. The pupillometry distribution (green) is centred about a substantially higher suppressive weight than for the other data types (note the logarithmic x-axis). The black curve shows the (scaled) prior distribution for the weight parameter.}\label{fig:modelfigure}
\end{figure}

@@

@@

@@

We conducted a conceptual replication of Experiment 1 using an alternative system. A pair of Spectra Tune Lab multiprimary devices (LEDmotive Technologies LLC, Barcelona, Spain) were coupled to a binocular headset using liquid light guides. The light was imaged onto a circular diffuser for each eye (field size 30 deg) with the central 8 degrees masked off using a black occluder. Therefore the replication experiment involved peripheral stimulation, unlike the main experiment which stimulated the central \textasciitilde4 degrees of the visual field. All conditions were otherwise as described for the main experiment, and we tested 12 participants in total.

>>>>

The pupillary light reflex is an automatic constriction of the iris sphincter muscles in response to increases in light levels, which causes the pupil to shrink (\protect\hyperlink{ref-McDougal2008}{McDougal and Gamlin, 2008}). There is a clear binocular component to this reflex, as stimulation of one eye still causes constriction of the other eye's pupil (termed the consensual response; Wyatt and Musselman (\protect\hyperlink{ref-Wyatt1981}{1981})). Importantly, the neuroanatomical pathway involved completely bypasses the canonical cortical pathway (retina to V1), instead involving a network of subcortical nuclei, including the Pretectal Olivary nucleus, Superior Cervical ganglion, and Edinger-Westphal nucleus (\protect\hyperlink{ref-Angee2021}{Angée et al., 2021}; \protect\hyperlink{ref-Mathot2018}{Mathôt, 2018}; \protect\hyperlink{ref-McDougal2008}{McDougal and Gamlin, 2008}; \protect\hyperlink{ref-Wang2015}{Wang and Munoz, 2015}). To account for the consensual response, these brain regions must combine information from the left and right eyes (\protect\hyperlink{ref-Doesschate1967}{Doesschate and Alpern, 1967}), yet the computation that achieves this is unclear. The pupil response can be modulated by periodic changes in luminance, and is temporally low-pass (\protect\hyperlink{ref-Barrionuevo2014}{Barrionuevo et al., 2014}; \protect\hyperlink{ref-Spitschan2014}{Spitschan et al., 2014}), most likely due to the mechanical limitations of the iris sphincter and dilator muscles (\protect\hyperlink{ref-Privitera2006}{Privitera and Stark, 2006}).

@@

We fitted a computational model to the data from Experiments 1 \& 3 using a hierarchical Bayesian approach. The model behaviour is displayed in Figure \ref{fig:modelfigure}a-d, with empirical data superimposed for comparison. In general, the model captures the key characteristics of the empirical data, with group-level parameter estimates provided in Table \ref{tab:paramtable}. We were particularly interested in comparing the weight of interocular suppression across data sets. We therefore plot the posterior distributions for this parameter for all four data sets (see Figure \ref{fig:modelfigure}e). The key finding is that the pupillometry results (green distribution) display a much greater weight of interocular suppression compared with the other data sets (grey, purple and yellow distributions). There is no overlap between the pupillometry distribution and any of the other three. All four distributions are also meaningfully below a weight of 1 -- the value that previous work using grating stimuli would predict (\protect\hyperlink{ref-Baker2017}{Baker and Wade, 2017}; \protect\hyperlink{ref-Meese2006}{Meese et al., 2006}), and the peak location of our prior distribution (black curve). These results offer an explanation of the empirical data: the strong interocular suppression for the pupillometry data is consistent with the weak binocular facilitation, and measurable dichoptic masking observed using that method. The weaker suppression for the other experiments is consistent with the near-linear binocular facilitation effects, and absent dichoptic masking.

@@

\caption{Summary of computational modelling. Panels (a-d) show empirical data from key conditions, replotted from earlier figures for the pupillometry (a), first harmonic EEG responses (b), second harmonic EEG responses (c) and contrast matching (d) experiments, with curves showing model behaviour generated using the median group-level parameter values.  Panel (e) shows the posterior probability distributions of the interocular suppression parameter for each of the four model fits. The pupillometry distribution (green) is centred about a substantially higher suppressive weight than for the other data types (note the logarithmic x-axis). The black curve shows the (scaled) prior distribution for the weight parameter.}\label{fig:modelfigure}
\end{figure}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Martin2023}{}}%
Martin JT, Boynton GM, Baker DH, Wade AR, Spitschan M. 2023. PySilSub: An open-source python toolbox for implementing the method of silent substitution in vision and non-visual photoreception research. \emph{Journal of Vision} \textbf{In press}.

\leavevmode\vadjust pre{\hypertarget{ref-Martin2022}{}}%
Martin JT, Pinto J, Bulte D, Spitschan M. 2022. PyPlr: A versatile, integrated system of hardware and software for researching the human pupillary light reflex. \emph{Behav Res Methods} \textbf{54}:2720--2739. doi:\href{https://doi.org/10.3758/s13428-021-01759-3}{10.3758/s13428-021-01759-3}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Privitera2006}{}}%
Privitera CM, Stark LW. 2006. A binocular pupil model for simulation of relative afferent pupil defects and the swinging flashlight test. \emph{Biol Cybern} \textbf{94}:215--24. doi:\href{https://doi.org/10.1007/s00422-005-0042-8}{10.1007/s00422-005-0042-8}

@@

We conducted a conceptual replication of Experiment 1 using an alternative system of hardware and software (\protect\hyperlink{ref-Martin2023}{Martin et al., 2023}, \protect\hyperlink{ref-Martin2022}{2022}). A pair of Spectra Tune Lab multiprimary devices (LEDmotive Technologies LLC, Barcelona, Spain) were coupled to a binocular headset using liquid light guides. The light was imaged onto a circular diffuser for each eye (field size 30 deg) with the central 8 degrees masked off using a black occluder. Therefore the replication experiment involved peripheral stimulation, unlike the main experiment which stimulated the central \textasciitilde4 degrees of the visual field. All conditions were otherwise as described for the main experiment, and we tested 12 participants in total.
